Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-11-20 ms/review-responses.tex

\begin{point}{}
    When neighbourhood size is small (10-100), the mean number of IBS tracts > 2bp (nIBS as in Table S1) is elevated similar to Wright's inbreeding coefficient, but mean of the distribution of pairwise IBS (mean(IBS)) is decreased. What could be the source of this discrepancy? How exactly mean(IBS) was calculated?
\end{point}

\reply{
The mean IBS is simply the mean of the vector of lengths of IBS tracts. 
The short answer is that if one splits a chromosome up into more pieces (increasing nIBS),
    the mean length of those pieces must necessarily be smaller.
    At low neighborhood sizes, spatially distant individuals tend to have longer coalescence times,
    leading to more (and hence smaller) IBS tracts.
    This can be seen in Supplemental Figure \ref{fig:allsumstats}.
    Note that skew(IBS) and number of long ($>$1e6) IBS tracts reflect other aspects of the distribution.

### 2019-11-20 ms/spaceness.tex

%\documentclass{article}
\newcommand{\beginsupplement}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }
\newcommand{\stopsupplement}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{\arabic{figure}}%
     }
\newcommand{\beginappendix}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{A\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{A\arabic{figure}}%
     }
\newcommand{\stopappendix}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{\arabic{figure}}%
     }

\documentclass[10pt,twoside,lineno,hidelinks]{preprint}
\usepackage{amsmath}
\usepackage[nodisplayskipstretch]{setspace}
\usepackage{makecell} 
\usepackage{changepage}
\usepackage{afterpage}
\setstretch{1}
\usepackage{placeins}
\NeedsTeXFormat{LaTeX2e}
\newcommand*{\articletypename}{PREPRINT}
% Use the documentclass option 'lineno' to view line numbers

@@

\subsection{Where are natural populations on this spectrum?}
For how much of the tree of life do spatial patterns circumscribe genomic variation? In Table \ref{table:NStable} we gathered estimates of neighborhood size from a range of organisms to get an idea of how likely dispersal is to play an important role in patterns of variation. These values should be compared to our simulation results with some caution, as though we expect neighborhood size to have similar effects across species of varying global $N_e$ \citep{Wright1946}, in our study we evaluated only a relatively small population of $\approx 10,000$ \llabel{popsizecaveat}. In addition, these empirical examples are likely biased towards small-neighborhood species (because few studies have quantified neighborhood size in species with very high dispersal or population density). However, from the available data we find that neighborhood sizes in the range we simulated are fairly common across a range of taxa. At the extreme low end of empirical neighborhood size estimates we see some flowering plants, large mammals, and colonial insects like ants. Species such as this have neighborhood size estimates small enough that spatial processes are likely to strongly influence inference. These include some human populations such as the Gainj- and Kalam-speaking people of Papua New Guinea, in which the estimated neighborhood sizes in \citep{Rousset1997} range from 40 to 410 depending on the method of estimation. Many more species occur in a middle range of neighborhood sizes between 100 and 1000 -- a range in which spatial processes play a minor role in our analyses under random spatial sampling but are important when sampling of individuals in space is clustered. Surprisingly, even some flying insects with huge census population sizes fall in this group, including fruit flies (\textit{D. melanogaster}) and mosquitoes (\textit{A. aegypti}). Last, many species likely have neighborhood sizes much larger than we simulated, including modern humans in northeastern Europe \citep{Ringbauer2017}. For these species demographic inference and summary statistics are likely to reflect minimal bias from spatial effects as long as dispersal is truly continuous across the landscape. While that is so we caution that association studies in which the effects of population structure are confounded with spatial variation in the environment are still sensitive to dispersal even at these large neighborhood sizes.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Supplement %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\beginappendix
\section{Appendix 1}
\llabel{appendix1}
%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Comparisons with Stepping-Stone Models}
We checked that our model produces reasonable results by comparing it to a reverse-time stepping-stone model implemented in msprime \citep{Kelleher2016}. In this class of models we imagine an $n \times n$ grid of populations exchanging migrants with neighboring populations at rate $m$. If these models are good approximations of the continuous case we expect that results will converge as $n \to \infty$, so we ran simulations while varying $n$ from 5 to 50 (Table \ref{table:grid_params}). To compare with continuous models we first distributed the the same "effective" number of individuals across the landscape as in our continuous-space simulations ($\approx 6100$, estimated from $\theta_{\pi}$ of random-mating continuous-space simulations). We then approximate the mean per-generation dispersal distance $\sigma$ given a total landscape width $W$ as the product of the probability of an individual being a migrant and the distance traveled by migrants: $\sigma \approx 4m(W/n)$. We ran 500 simulations per $n$ while sampling $\sigma$ from $U(0.2,4)$. We then randomly selected 60 diploid individuals from each simulation (approximating diploidy by combining pairs of chromosomes with contiguous indices within demes) and calculated a set of six summary statistics using the scripts described in the summary statistics portion of the main text.

@@

In general we find many of the qualitative trends are similar among continuous and stepping-stone models and that, in most cases, statistics from stepping-stone models approach the continuous model as the resolution of the grid increases. For example, $\theta_\pi$ is inflated at low neighborhood sizes (i.e. low $m$), and the extent of the inflation increases to approach the continuous case as the resolution of the landscape increases. Similar pattterns are observed for $F_{is}$ and observed heterozygosity. However, $\theta_{W}$ behaves differently, with increased grid resolution leading to lower values. This in turn drives an even more positive Tajima's $D$ in grid simulations at small neighborhood sizes.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Demographic model}
\label{apx:demographic_model}
\llabel{dem_model_app}

@@

\stopappendix
\clearpage

\beginsupplement
\section{Supplementary Figures and Tables}
\FloatBarrier

@@

\afterpage{\clearpage}
\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/sumstats_by_neighbors_allstats.pdf}
\caption{Change in summary statistics by neighborhood size and sampling scheme calculated from simulated sequence data of 60 individuals.}
\label{fig:allsumstats} 
\end{figure*}
\FloatBarrier

@@

\afterpage{\clearpage}
\begin{figure*}[p]
\centering
\includegraphics[height=0.9\textheight]{figures/manhattan_plots_combo-01.png}
\caption{Manhattan plots for a sample of simulations at varying neighborhood sizes. Labels on the right of each plot describes the spatial distribution of environmental factors (described in the methods section of the main text). Points in red are significantly associated with a nongenetic phenotype using a 5\% FDR threshold (dashed line). For runs with significant associations the dotted line is a Bonferroni-adjusted cutoff for $p=0.05$.}
\label{fig:manhattan_plots}
\end{figure*}
\FloatBarrier

\begin{table}
\small
\centering
\caption{\bf Summary statistics calculated on simulated genotypes.}
\begin{tabular}{rllrrrrr}
  \hline
Statistic & Description \\ 
  \hline
$\Theta_{/pi}$ & Mean of the distribution of pairwise genetic differences \\
$\Theta_{/W}$ & Effective population size based on segregating sites \\
Segregating Sites & Total number of segregating sites in the sample \\
Tajima's $D$ & Difference in $\Theta_{/pi}$ and $\Theta_{/W}$ over its standard deviation\\
Observed Heterzygosity & Proportion of heterozygous individuals in the sample \\
$F_{IS}$ & Wright's inbreeding coefficient $1-H_{e}/H_{o}$ \\
$var(D_{xy})$ & Variance in the distribution of pairwise genetic distances \\
$skew(D_{xy})$ & Skew of the distribution of pairwise genetic distances \\
$mean(IBS)$ & \makecell[l]{Mean of the distribution of pairwise identical-by-state (IBS) \\tract lengths taken over all pairs.} \\
$var(IBS)$ & \makecell[l]{Variance of the distribution of pairwise identical-by-state (IBS) \\tract lengths taken over all pairs.} \\
$skew(IBS)$ & \makecell[l]{Skew of the distribution of pairwise identical-by-state (IBS) \\tract lengths taken over all pairs.} \\
$nIBS$ & Mean number of IBS tracts with length > 2bp across all pairs in the sample. \\
$nIBS>1e6$ & \makecell[l]{Mean number of IBS tracts over $1\times10^6$bp per pair \\across all pairs in the sample.} \\ 
$corr(D_{xy},dist)$ & Pearson correlation between genetic distance and $\log_{10}(spatial\text{ }distance)$ \\
$corr(mean(IBS),dist)$ & \makecell[l]{Pearson correlation between the mean of the IBS tract distribution \\for each pair of samples and $\log_{10}(spatial\text{ }distance)$} \\
$corr(nIBS,dist)$ & \makecell[l]{Pearson correlation between the number of IBS tracts \\for each pair of samples and $\log_{10}(spatial\text{ }distance)$} \\
$corr(IBS>1e6,dist)$ & \makecell[l]{Pearson correlation between the number of IBS tracts $> 1\times10^6$bp \\for each pair of samples and $\log_{10}(spatial\text{ }distance)$} \\
$corr(skew(IBS),dist)$ & \makecell[l]{Pearson correlation between the skew of the distribution of pairwise haplotype\\ block lengths for each pair of samples and $\log_{10}(spatial\text{ }distance)$} \\
\end{tabular}
\label{table:sumstats}
\end{table}
\FloatBarrier

@@

\stopsupplement

>>>>

%\documentclass{article}
\documentclass[10pt,twoside,lineno,hidelinks]{preprint}
\usepackage{amsmath}
\usepackage[nodisplayskipstretch]{setspace}
\usepackage{makecell} 
\usepackage{changepage}
\usepackage{afterpage}
\setstretch{1}
\usepackage{placeins}
\NeedsTeXFormat{LaTeX2e}
\newcommand*{\articletypename}{PREPRINT}
% Use the documentclass option 'lineno' to view line numbers

@@

\subsection{Where are natural populations on this spectrum?}
For how much of the tree of life do spatial patterns circumscribe genomic variation? In Table \ref{table:NStable} we gathered estimates of neighborhood size from a range of organisms to get an idea of how likely dispersal is to play an important role in patterns of variation. These values should be compared to our simulation results with some caution, as though we expect neighborhood size to have similar effects across species of varying global $N_e$ \citep{Wright1946}, in our study we evaluated only a relatively small population of $\approx 10,000$. \llabel{popsizecaveat}
In addition, these empirical examples are likely biased towards small-neighborhood species (because few studies have quantified neighborhood size in species with very high dispersal or population density). However, from the available data we find that neighborhood sizes in the range we simulated are fairly common across a range of taxa. At the extreme low end of empirical neighborhood size estimates we see some flowering plants, large mammals, and colonial insects like ants. Species such as this have neighborhood size estimates small enough that spatial processes are likely to strongly influence inference. These include some human populations such as the Gainj- and Kalam-speaking people of Papua New Guinea, in which the estimated neighborhood sizes in \citep{Rousset1997} range from 40 to 410 depending on the method of estimation. Many more species occur in a middle range of neighborhood sizes between 100 and 1000 -- a range in which spatial processes play a minor role in our analyses under random spatial sampling but are important when sampling of individuals in space is clustered. Surprisingly, even some flying insects with huge census population sizes fall in this group, including fruit flies (\textit{D.~melanogaster}) and mosquitoes (\textit{A. aegypti}). Last, many species likely have neighborhood sizes much larger than we simulated, including modern humans in northeastern Europe \citep{Ringbauer2017}. For these species demographic inference and summary statistics are likely to reflect minimal bias from spatial effects as long as dispersal is truly continuous across the landscape. While that is so we caution that association studies in which the effects of population structure are confounded with spatial variation in the environment are still sensitive to dispersal even at these large neighborhood sizes.

%%%%%%%%%%%
\subsection{Other demographic models}

Any simulation of a population of reproducing organisms for a reasonably long amount of time
requires some kind of control on population sizes,
or else the population will either die out or grow very large. \llabel{demog_disco}
The usual choice of population regulation for population genetics --
a constant size, as in the Wright--Fisher model --
implies biologically unrealistic interactions between geographically distant parts of the species range.
Our choice to regulate population size by including a local density-dependent control on mortality
is only one of many possible ways to do this.
We could have instead regulated fecundity, or recruitment, or both;
this general class of models
is sometimes referred to as the ``Bolker--Pacala model'' \citep{bolker1997using}.
It is not currently clear how much different choices of vital rates,
or of functional forms for the regulation,
might quantitatively affect our results,
although the general predictions should be robust to similar forms of regulation.
Since populations are still entirely \emph{intrinsically} regulated,
our model still has a very strong ``population genetics'' flavor.
Alternatively, population size could be regulated by interactions with other species
(e.g., a Lotka-Volterra model),
or extrinsically specified by local resource availability
(e.g., by food or nest site availability).
Indeed, our model could be interpreted as a caricature of such a model:
as local density increases, good habitat is increasingly occupied,
pushing individuals into more marginal habitat and increasing their mortality.
Many such models should be behave similarly to ours,
but others (especially those with local population cycling), may differ dramatically.
There is a clear need for greater exploration of 
the consequences for population genetics of ecologically realistic population models.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Supplement %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \beginappendix
\appendix
\setcounter{table}{0}
\renewcommand{\thetable}{A\arabic{table}}%
\setcounter{figure}{0}
\renewcommand{\thefigure}{A\arabic{figure}}%

%%%%%%%%%%%%%%%%%%%%%%%
\section{Comparisons with Stepping-Stone Models}
\llabel{apx:stepping_stone}

We checked that our model produces reasonable results by comparing it to a reverse-time stepping-stone model implemented in msprime \citep{Kelleher2016}. In this class of models we imagine an $n \times n$ grid of populations exchanging migrants with neighboring populations at rate $m$. If these models are good approximations of the continuous case we expect that results will converge as $n \to \infty$, so we ran simulations while varying $n$ from 5 to 50 (Table \ref{table:grid_params}). To compare with continuous models we first distributed the the same "effective" number of individuals across the landscape as in our continuous-space simulations ($\approx 6100$, estimated from $\theta_{\pi}$ of random-mating continuous-space simulations). We then approximate the mean per-generation dispersal distance $\sigma$ given a total landscape width $W$ as the product of the probability of an individual being a migrant and the distance traveled by migrants: $\sigma \approx 4m(W/n)$. We ran 500 simulations per $n$ while sampling $\sigma$ from $U(0.2,4)$. We then randomly selected 60 diploid individuals from each simulation (approximating diploidy by combining pairs of chromosomes with contiguous indices within demes) and calculated a set of six summary statistics using the scripts described in the summary statistics portion of the main text.

@@

In general we find many of the qualitative trends are similar among continuous and stepping-stone models and that, in most cases, statistics from stepping-stone models approach the continuous model as the resolution of the grid increases. For example, $\theta_\pi$ is inflated at low neighborhood sizes (i.e. low $m$), and the extent of the inflation increases to approach the continuous case as the resolution of the landscape increases. Similar patterns are observed for $F_{is}$ and observed heterozygosity. However, $\theta_{W}$ behaves differently, with increased grid resolution leading to lower values. This in turn drives an even more positive Tajima's $D$ in grid simulations at small neighborhood sizes.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Demographic model}
\llabel{apx:demographic_model}

@@

\clearpage

% \beginsupplement
\setcounter{table}{0}
\renewcommand{\thetable}{S\arabic{table}}%
\setcounter{figure}{0}
\renewcommand{\thefigure}{S\arabic{figure}}%

\section{Supplementary Figures and Tables}
\FloatBarrier

@@

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/sumstats_by_neighbors_allstats.pdf}
\caption{Change in summary statistics by neighborhood size and sampling scheme calculated from simulated sequence data of 60 individuals.}
\label{fig:allsumstats} 
\end{figure*}
\FloatBarrier

@@

\begin{figure*}[p]
\centering
\includegraphics[height=0.9\textheight]{figures/manhattan_plots_combo-01.png}
\caption{Manhattan plots for a sample of simulations at varying neighborhood sizes. Labels on the right of each plot describes the spatial distribution of environmental factors (described in the methods section of the main text). Points in red are significantly associated with a nongenetic phenotype using a 5\% FDR threshold (dashed line). For runs with significant associations the dotted line is a Bonferroni-adjusted cutoff for $p=0.05$.}
\label{fig:manhattan_plots}
\end{figure*}
\FloatBarrier

\begin{table}
\small
\centering
\caption{\bf Summary statistics calculated on simulated genotypes.}
\begin{tabular}{rllrrrrr}
  \hline
Statistic & Description \\ 
  \hline
$\Theta_{/pi}$ & Mean of the distribution of pairwise genetic differences \\
$\Theta_{/W}$ & Effective population size based on segregating sites \\
Segregating Sites & Total number of segregating sites in the sample \\
Tajima's $D$ & Difference in $\Theta_{/pi}$ and $\Theta_{/W}$ over its standard deviation\\
Observed Heterozygosity & Proportion of heterozygous individuals in the sample \\
$F_{IS}$ & Wright's inbreeding coefficient $1-H_{e}/H_{o}$ \\
$var(D_{xy})$ & Variance in the distribution of pairwise genetic distances \\
$skew(D_{xy})$ & Skew of the distribution of pairwise genetic distances \\
$mean(IBS)$ & \makecell[l]{Mean of the distribution of pairwise identical-by-state (IBS) \\tract lengths taken over all pairs.} \\
$var(IBS)$ & \makecell[l]{Variance of the distribution of pairwise identical-by-state (IBS) \\tract lengths taken over all pairs.} \\
$skew(IBS)$ & \makecell[l]{Skew of the distribution of pairwise identical-by-state (IBS) \\tract lengths taken over all pairs.} \\
$nIBS$ & Mean number of IBS tracts with length > 2bp across all pairs in the sample. \\
$nIBS>1e6$ & \makecell[l]{Mean number of IBS tracts over $1\times10^6$bp per pair \\across all pairs in the sample.} \\ 
$corr(D_{xy},dist)$ & Pearson correlation between genetic distance and $\log_{10}(spatial\text{ }distance)$ \\
$corr(mean(IBS),dist)$ & \makecell[l]{Pearson correlation between the mean of the IBS tract distribution \\for each pair of samples and $\log_{10}(spatial\text{ }distance)$} \\
$corr(nIBS,dist)$ & \makecell[l]{Pearson correlation between the number of IBS tracts \\for each pair of samples and $\log_{10}(spatial\text{ }distance)$} \\
$corr(IBS>1e6,dist)$ & \makecell[l]{Pearson correlation between the number of IBS tracts $> 1\times10^6$bp \\for each pair of samples and $\log_{10}(spatial\text{ }distance)$} \\
$corr(skew(IBS),dist)$ & \makecell[l]{Pearson correlation between the skew of the distribution of pairwise haplotype\\ block lengths for each pair of samples and $\log_{10}(spatial\text{ }distance)$} \\
\end{tabular}
\label{table:sumstats}
\end{table}
\FloatBarrier

@@
